\bf Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-07-11 manuscript/main.tex
proposition 1 and 2

	We now describe the strategy for computing the PLS component direction at the $l$-th iteration. For simplicity, we omit the iteration index and assume that the observations have been residualized and are orthogonal to those from previous steps (see Section~\ref{subsec: Step 2: Computing Predictors via Hybrid-on-Scalar Regression} for the residualization procedure).
	The goal is to find a unit-norm direction $\xi \in 
	(\mathcal{B}^K \times \mathbb{R}^p) \subset \mathcal{H}$ that maximizes the squared empirical covariance between the resulting PLS component scores $\langle \widetilde{W}_1, \xi \rangle_{\mathcal{H}}, \ldots, \langle \widetilde{W}_n, \xi \rangle_{\mathcal{H}}$ and the responses $y_1, \ldots, y_n$,
	 defined as
	 		\begin{equation}\label{def:squared_empirical_cov}
	 	\widehat{\textnormal{Cov}}^2 \left( \langle \widetilde{\mathbf{W}}, \xi \rangle_{\mathcal{H}}, Y \right)
	 	:=
	 	\biggl(
	 	\frac{1}{n} 
	 	\sum_{i=1}^n
	 	y_i
	 	\langle \widetilde{W}_i, \xi \rangle_{\mathcal{H}}
	 	\biggr)^2
	 	.
	 \end{equation}
	We denote the maximizer as
	\begin{equation*}
		\hat{\xi} 
= 
(\hat{\xi}_1, \ldots, \hat{\xi}_K, \hat{\boldsymbol{\zeta}})
:= \arg \max_{\xi \in (\mathcal{B}^K \times \mathbb{R}^p) \subset \mathcal{H}} \widehat{\textnormal{Cov}}^2 \left( \langle \widetilde{\mathbf{W}}, \xi \rangle_{\mathcal{H}}, Y \right)
	\end{equation*}
%
	where each functional part $\hat{\xi}_j $  is expressed as
	\begin{equation*}
		\hat{\xi}_j(t) := 
		\sum_{m=1}^M \hat{d}_{jm} \, b_m(t), \quad j = 1, \ldots, K.
	\end{equation*}

@@

	\begin{proposition}\label{proposition:generalized_rayleigh_quotient}
At the $l$-th iteration of the PLS algorithm, the unit-norm direction in $(\mathcal{B}^K \times \mathbb{R}^p) \subset \mathcal{H}$ that maximizes the squared empirical covariance \eqref{def:squared_empirical_cov} is the solution to the following generalized eigenvalue problem:
\begin{align}
	\left(
	\hat{\mathbf{d}}_1^\top, \ldots, \hat{\mathbf{d}}_K^\top, \hat{\boldsymbol{\zeta}}^\top
	\right)^\top
	=
	\arg \max_{\boldsymbol{\xi} \in \mathbb{R}^{MK+p}}   \boldsymbol{\xi}^\top V \boldsymbol{\xi}
	\quad \text{subject to} \quad \boldsymbol{\xi}^\top B^\ast \boldsymbol{\xi} = 1,
	\label{eq: Regularized generalized rayleigh quotient equation}
\end{align}
where
$$
V := \frac{1}{n^2} (B^\ast \Theta^\top \mathbf{y})(B^\ast \Theta^\top \mathbf{y})^\top \in \mathbb{R}^{(MK+p) \times (MK+p)},
$$
where
$B^\ast := \operatorname{blkdiag}(B, \ldots, B, I_p)  \in  \mathbb{R}^{(MK+p) \times (MK+p)}$,
 $\Theta := (\Theta_1, \ldots, \Theta_K, Z) \in \mathbb{R}^{n \times (MK + p)}$,
$B \in \mathbb{R}^{M \times M}$ is the gram matrix  of the basis functions  with entries
$
B_{m, m'} := \langle b_m, b_{m'} \rangle = \int_0^1  b_m(t)  b_{m'}(t) dt
$
for $m, m' = 1, \ldots, M$.
\end{proposition}
Since $V$ is a rank 1 matrix, the solution of the generalized eigenvalue problem is easily obtained by matrix multiplication and normalization:
	\begin{proposition} \label{proposition:generalized_rayleigh_quotient_linear}
	We can compute  $\hat{\mathbf{d}}_1, \ldots, \hat{\mathbf{d}}_K$ and $\hat{\boldsymbol{\zeta}}$ in Proposition \ref{proposition:generalized_rayleigh_quotient} as 
\begin{equation*}
	\hat{\mathbf{d}}_j  = \frac{1}{\sqrt{\sum_{j=1}^K 
			\mathbf{y}^\top
			\Theta_j
			B \Theta_j^\top \mathbf{y} 
			+ 
			\mathbf{y}^\top Z Z^\top \mathbf{y}}}
	\Theta_j^\top \mathbf{y},
	\quad
	%%%%%%%%%%%%%%%%
	\hat{\boldsymbol{\zeta}}  = \frac{1}{\sqrt{\sum_{j=1}^K 
			\mathbf{y}^\top
			\Theta_j
			B \Theta_j^\top \mathbf{y} 
			+ 
			\mathbf{y}^\top Z Z^\top \mathbf{y}}}
	Z^\top \mathbf{y}.
\end{equation*}
\end{proposition}
The expressions above represent matrix–vector multiplications normalized by a common factor. 
Proofs of Propositions \ref{proposition:generalized_rayleigh_quotient} and \ref{proposition:generalized_rayleigh_quotient_linear} are provided in Appendix~\ref{section:proof:proposition:generalized_rayleigh_quotient_linear_system_unregularized}.
Although the unnormalized coefficients are computed separately for the functional and scalar part of the PLS component direction, the normalization step couples them, allowing the components to influence each other. This enables the procedure to account for the correlation between functional and scalar components.

>>>>

	We now describe the strategy for computing the PLS component direction at the $l$-th iteration. For simplicity, we omit the iteration index and assume that the observations have been residualized and are orthogonal to those from previous steps (see Section~\ref{subsec: Step 2: Computing Predictors via Hybrid-on-Scalar Regression} for the residualization procedure).
	The goal is to find a unit-norm direction $\xi \in 
	(\mathcal{B}^K \times \mathbb{R}^p) \subset \mathcal{H}$ that maximizes the squared empirical covariance between the resulting PLS component scores $\langle \widetilde{W}_1, \xi \rangle_{\mathcal{H}}, \ldots, \langle \widetilde{W}_n, \xi \rangle_{\mathcal{H}}$ and the responses $y_1, \ldots, y_n$,
	 defined as
\begin{equation}\label{def:squared_empirical_cov}
	 	\widehat{\textnormal{Cov}}^2 \left( \langle \widetilde{\mathbf{W}}, \xi \rangle_{\mathcal{H}}, Y \right)
	 	:=
	 	\biggl(
	 	\frac{1}{n} 
	 	\sum_{i=1}^n
	 	y_i
	 	\langle \widetilde{W}_i, \xi \rangle_{\mathcal{H}}
	 	\biggr)^2
	 	.
	 \end{equation}
	We denote the maximizer as
	\begin{equation}\label{def:maximizer_squared_empirical_cov}
		\hat{\xi} 
= 
(\hat{\xi}_1, \ldots, \hat{\xi}_K, \hat{\boldsymbol{\zeta}})
:= \arg \max_{\xi \in (\mathcal{B}^K \times \mathbb{R}^p) \subset \mathcal{H}} \widehat{\textnormal{Cov}}^2 \left( \langle \widetilde{\mathbf{W}}, \xi \rangle_{\mathcal{H}}, Y \right)
	\end{equation}
%
	where each functional part $\hat{\xi}_j $  is expressed as
	\begin{equation*}
		\hat{\xi}_j(t) := 
		\sum_{m=1}^M \hat{d}_{jm} \, b_m(t), \quad j = 1, \ldots, K.
	\end{equation*}

@@

	\begin{proposition}\label{proposition:generalized_rayleigh_quotient}
At the $l$-th iteration of the PLS algorithm, let $
( \hat{\mathbf{d}}_1^\top, \ldots, \hat{\mathbf{d}}_K^\top, \hat{\boldsymbol{\zeta}}^\top )^\top
$ denote the coefficients 
of $	\hat{\xi}$, 
defined in \eqref{def:maximizer_squared_empirical_cov},
which is
the projected unit-norm direction that maximizes the squared empirical covariance. These coefficients can be obtained by solving the following generalized eigenvalue problem
\begin{align}
	\left(
	\hat{\mathbf{d}}_1^\top, \ldots, \hat{\mathbf{d}}_K^\top, \hat{\boldsymbol{\zeta}}^\top
	\right)^\top
	=
	\arg \max_{\boldsymbol{\xi} \in \mathbb{R}^{MK+p}}   \boldsymbol{\xi}^\top V \boldsymbol{\xi}
	\quad \text{subject to} \quad \boldsymbol{\xi}^\top B^\ast \boldsymbol{\xi} = 1,
	\label{eq: Regularized generalized rayleigh quotient equation}
\end{align}
where
$$
V := \frac{1}{n^2} (B^\ast \Theta^\top \mathbf{y})(B^\ast \Theta^\top \mathbf{y})^\top \in \mathbb{R}^{(MK+p) \times (MK+p)},
$$
where
$B^\ast := \operatorname{blkdiag}(B, \ldots, B, I_p)  \in  \mathbb{R}^{(MK+p) \times (MK+p)}$,
 $\Theta := (\Theta_1, \ldots, \Theta_K, Z) \in \mathbb{R}^{n \times (MK + p)}$,
$B \in \mathbb{R}^{M \times M}$ is the gram matrix  of the basis functions  with entries
$
B_{m, m'} := \langle b_m, b_{m'} \rangle = \int_0^1  b_m(t)  b_{m'}(t) dt
$
for $m, m' = 1, \ldots, M$.
\end{proposition}
Since $V$ is a rank 1 matrix, the solution of the generalized eigenvalue problem is easily obtained by matrix multiplication and normalization:
	\begin{proposition} \label{proposition:generalized_rayleigh_quotient_linear}
	We can compute  $\hat{\mathbf{d}}_1, \ldots, \hat{\mathbf{d}}_K$ and $\hat{\boldsymbol{\zeta}}$ in Proposition \ref{proposition:generalized_rayleigh_quotient} as 
\begin{equation*}
	\hat{\mathbf{d}}_j  = \frac{1}{\sqrt{\sum_{j=1}^K 
			\mathbf{y}^\top
			\Theta_j
			B \Theta_j^\top \mathbf{y} 
			+ 
			\mathbf{y}^\top Z Z^\top \mathbf{y}}}
	\Theta_j^\top \mathbf{y},
	\quad
	%%%%%%%%%%%%%%%%
	\hat{\boldsymbol{\zeta}}  = \frac{1}{\sqrt{\sum_{j=1}^K 
			\mathbf{y}^\top
			\Theta_j
			B \Theta_j^\top \mathbf{y} 
			+ 
			\mathbf{y}^\top Z Z^\top \mathbf{y}}}
	Z^\top \mathbf{y}.
\end{equation*}
\end{proposition}
The expressions above represent matrix–vector multiplications normalized by a common factor. 
Proofs of Propositions \ref{proposition:generalized_rayleigh_quotient} and \ref{proposition:generalized_rayleigh_quotient_linear} are provided in Appendix~\ref{section:proof:proposition:generalized_rayleigh_quotient_linear_system_unregularized}.
Although the unnormalized coefficients are computed separately for the functional and scalar part of the PLS component direction, the normalization step couples them, allowing the components to influence each other. This enables the procedure to account for the correlation between functional and scalar components.
